---
title: "Coursera-Getting and Cleaning Data-Week 3"
author: "Zijun Liu"
date: "`r Sys.time()`"
output:
  html_document:
    highlight: haddock
    theme: default
    code_folding: hide
    df_print: paged
    toc: yes
    toc_depth: '3'
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---


# Subsetting and Sorting

## Subsetting and Sorting
```{r create dataset}
set.seed(13435)
X <- data.frame("var1" = sample(1:5), "var2" = sample(6:10), "var3" = sample(11:15))
X <- X[sample(1:5), ]; X$var2[c(1, 3)] <- NA
X

```

```{r subsetting quick view}
# Subset column using index
X[, 1]

# Subset column using column name
X[, "var1"]


# Subset both row and column
X[1:2, "var1"]

# Subset using logicals
X[(X$var1 <= 3 & X$var3 > 11), ]
X[(X$var1 <= 3 | X$var3 > 11), ]

# Deal with missing values
X[X$var2 > 8, ] # include missing var2
X[which(X$var2 > 8), ] # not include missing var2
```

```{r sorting}
sort(X$var1)
sort(X$var1, decreasing = TRUE)
sort(X$var2, na.last = TRUE)
```

## Ordering and adding rows/columns
```{r ordering}

# reordering the rows so var1 is in increasing order
X[order(X$var1), ]

X[order(X$var1, X$var3), ] # if multiple same value of var1, sort by var3

# with tidyverse
library(tidyverse)
X %>% 
    arrange(var1)
```

```{r adding rows and columns}
X$var4 <- rnorm(5)
X

Y <- cbind(X, rnorm(5))
Y
```


# Summarizing Data

```{r Getting data from the web}
if(!file.exists("./data")){dir.create("./data")}
fileUrl <- "https://data.baltimorecity.gov/api/views/k5ry-ef3g/rows.csv?accessType=DOWNLOAD"
download.file(fileUrl, destfile="./data/restaurants.csv", method = "curl")
restData <- read.csv("./data/restaurants.csv")

```


```{r Summary of a dataset, results=FALSE}
head(restData, n=3)
tail(restData, n=3)
summary(restData)
str(restData)

quantile(restData$councilDistrict, na.rm = TRUE)
quantile(restData$councilDistrict, probs = c(0.5, 0.75, 0.9))

table(restData$zipCode, useNA = "ifany")
table(restData$councilDistrict, restData$zipCode)
```

```{r check for missing values, results=FALSE}
sum(is.na(restData$councilDistrict))
# [1] 0 
any(is.na(restData$councilDistrict))
# [1] FALSE

# if all zipcode are freater than 0
all(restData$zipCode > 0)

# row and column sums
is.na(restData)
colSums(is.na(restData))

all(colSums(is.na(restData)) == 0)
```

```{r values with specific characteristics}
table(restData$zipCode %in% c("21212", "21213"))

# subset the dataset
restData[restData$zipCode %in% c("21212", "21213"), ]
```

```{r cross tabs}
data(UCBAdmissions)
DF = as.data.frame(UCBAdmissions)
summary(DF)

xt <- xtabs(Freq ~ Gender + Admit, data = DF)
xt
```

```{r flat tables}
warpbreaks$replicate <- rep(1:9, len = 54)
xt = xtabs(breaks~., data = warpbreaks)
xt

ftable(xt)
```

```{r size of a data set}
fakeData = rnorm(1e5)
object.size(fakeData)
# 800048 bytes

print(object.size(fakeData), units = "Mb")
# 0.8 Mb
```

## Creating new variables

* Missingness indicators
* "Cutting up" quantitiative variables
* Applying transforms
```{r creating sequences}

s1 <- seq(1, 10, by = 2); s1
s2 <- seq(1, 10, length = 3); s2

# along: take the length from the length of this argument.
x <- c(1, 3, 8, 25, 100); seq(along = x) 

```

```{r create new variables}
# subsetting variables
restData$nearMe <- restData$neighborhood %in% c("Roland Park", "Homeland")
table(restData$nearMe)

# Creating binary variables
restData$zipWrong <- ifelse(restData$zipCode < 0, TRUE, FALSE)
table(restData$zipWrong)

# Creating categorical variables
restData$zipGroups <- cut(restData$zipCode, breaks = quantile(restData$zipCode))
table(restData$zipGroups)

table(restData$zipGroups, restData$zipCode)

# easier cutting
library(Hmisc)
restData$zipGroups = cut2(restData$zipCode, g = 4)
table(restData$zipGroups)

```


## Reshaping Data  
The goal is tidy data:
* Each variable forms a columns
* Each observation forms a row
* Each table/file stores data about one knid of observation (e.g. people/hospitals).

```{r the new pivot function}
relig_income
relig_income %>% 
    pivot_longer(-religion, names_to = "income", values_to = "count")

## average values
head(InsectSprays)
tapply(InsectSprays$count, InsectSprays$spray, sum)

## Another way: split, apply, then combine
spIns <- split(InsectSprays$count, InsectSprays$spray)
spIns
sprCount <- lapply(spIns, sum)
sprCount
unlist(sprCount)
# or
sapply(spIns, sum)
```

